\pkg{hesim}: Health Economic Simulation Modeling and Decision Analysis

### 2021-08-01 manuscript.Rnw
More edits to better compare R packages

The functionality of existing packages can generally be separated into three categories: estimation of model parameters, simulation of model outcomes, and decision analysis. Nearly all packages mentioned above support decision analysis in some capacity: \pkg{BCEA}, \pkg{SAVI}, \pkg{dampack}, \pkg{EVSI}, and \pkg{heemod} all contain features for post-processing the output of simulation models in order to quantify decision uncertainty and conduct value of information analyses. On the other hand, there is, to our knowledge, only one package, \pkg{heemod}, that is designed for simulation of health economic models. \pkg{heemod} supports development of Markov cohort models and can incorporate parameter uncertainty for a PSA. Finally, it is important to note that one of, it not the primary advantages of using \proglang{R} for cost-effectiveness modeling is the rich set of packages available for statistical analysis. While any number of these can be used for model parameter estimation---as will be partially demonstrated in this paper---\pkg{survHE} is unique in that it was designed specifically for use in health economic evaluations and consequently contains features such as convenient comparisons of multiple parametric survival models and simulation of model parameters for PSA.

@@

Of the existing packages, \pkg{hesim} is most similar to \pkg{heemod} in that it is designed primarily for simulation of economic models, but can also be used to conduct a CEA. That said, there are some notable differences. First, \pkg{hesim} can be used for individual-level simulations, whereas \pkg{heemod} only supports cohort-level models. Second, all outcomes in \pkg{hesim} are simulated across treatment strategies and patient subgroups within a PSA in a vectorized fashion. This not only increases speed because outcomes are simulated at the \proglang{C++} level, but also makes it easier to account for patient heterogeneity and compare a large number of treatment strategies. Third, since all model types in \pkg{hesim} can be parameterized by fitting statistical models in \proglang{R}, it is easier to integrate the statistical analysis and economic model structure.

@@

As mentioned in the introduction, \pkg{heemod} is the only other general purpose \proglang{R} package for cost-effectiveness modeling. Unlike \pkg{hesim}, simulations are not vectorized. To asses the significance of this, performance comparisons are provided on the package website. The time-inhomogeneous total hip replacement example from \citet{briggs2006decision} was run with both packages using 1,000 PSA iterations. At the time of writing, a cohort model in \pkg{heemod} completed in approximately 1.9 minutes while the same cohort model took approximately 1 second to run in \pkg{hesim} and an equivalent individual-level simulation in \pkg{hesim} ran in 7.4 seconds.

@@

\pkg{mstate} is another package for multi-state modeling that is designed for estimation of non-parametric and semi-parametric (i.e., Cox proportional hazards) models \citep{de2011mstate}. It supports prediction of state probabilities and simulation of multi-state trajectories (to facilitate prediction for clock-reset models) from cumulative hazard functions evaluated at specific times. \citet{williams2017cost} has adapted \pkg{mstate} for cost-effectiveness modeling applications. There are however, some disadvantages of using \pkg{mstate} for health economic modeling since it was not designed for it. \code{mssample()}---the core function for simulating multi-state models---does not natively support heterogeneous target populations, PSA, simultaneous simulation of multiple treatment strategies, or computation of QALYs and costs from the simulated output. It is also less computationally efficient than \pkg{hesim} for two main reasons. First, simulation of general cumulative hazards functions is considerably slower than simulation of parametric distributions when efficient random number generators are available. Second, it is written completely in \proglang{R} and incorporating PSA, multiple treatment strategies, or heterogeneous patients results in code that is not vectorized. To compare computational performance, we ran 1,000 PSA iterations where 1 treatment strategy and 1,000 homogeneous patients were simulated through a 6-state Weibull model during each iteration. The model completed in 2.8 seconds in \pkg{hesim} and 3.1 hours in \pkg{mstate}, showing that computational considerations are not just an academic concern.

A number of extensions of the current version of \pkg{hesim} would be helpful. First, we plan to extend the individual-level simulation so that users can update input data during the simulation as a function of time elapsed and/or the health states transitioned to. This would allow potentially time-varying variables such as age to update each time a patient transitions to a new state. Second, we would like to add more examples to ensure that \pkg{hesim} is fit for purpose for a wide range of applications in health economics. Third, more complex models of utility \citep{kharroubi2007modelling, hernandez2013relationship} and costs \citep{nixon2005methods} could help better capture heterogeneity in preferences and spending. Fourth, it would be useful to add functionality that automatically creates tunnel states for users in cDTSTMs. Fifth, equity considerations could be incorporated by adding support for distributional CEA \citep{asaria2015distributional, asaria2016distributional, cookson2017using}. Finally, a larger aspiration is to provide support for parameterization via NMA, particularly for estimation of relative treatment effects in survival (e.g., for PSMs) and multi-state models. However, since existing \proglang{R} packages are lacking in this area, this could involve creation of a separate NMA package and a new dependency for \pkg{hesim}.

>>>>

The functionality of existing packages can generally be separated into three categories: estimation of model parameters, simulation of model outcomes, and decision analysis. Nearly all packages mentioned above can be used to perform decision analysis in some capacity: \pkg{BCEA}, \pkg{SAVI}, \pkg{dampack}, \pkg{EVSI}, and \pkg{heemod} all contain features for post-processing the output of simulation models in order to quantify decision uncertainty and conduct value of information analyses. On the other hand, there is, to our knowledge, only one package, \pkg{heemod}, that is designed for simulation of health economic models. \pkg{heemod} can be used to develop Markov cohort models and can incorporate parameter uncertainty for a PSA. Finally, it is important to note that one of, it not the primary advantages of using \proglang{R} for cost-effectiveness modeling is the rich set of packages available for statistical analysis. While any number of these can be used for model parameter estimation---as will be partially demonstrated in this paper---\pkg{survHE} is unique in that it was designed specifically for use in health economic evaluations and consequently contains features such as convenient comparisons of multiple parametric survival models and simulation of model parameters for PSA.

@@

Of the existing packages, \pkg{hesim} is most similar to \pkg{heemod} in that it is designed primarily for simulation of economic models, but can also be used to conduct a CEA. That said, there are some notable differences. First, \pkg{hesim} can be used for individual-level simulations, whereas \pkg{heemod} can only simulate cohort-level models. Second, all outcomes in \pkg{hesim} are simulated across parameter samples for a PSA, treatment strategies, and patient subgroups in a vectorized fashion. This not only increases speed because outcomes are simulated at the \proglang{C++} level, but also makes it easier to account for patient heterogeneity and compare a large number of treatment strategies. Third, since all model types in \pkg{hesim} can be parameterized by fitting statistical models in \proglang{R}, it is easier to integrate the statistical analysis and economic model structure.

@@

@@

\pkg{mstate} is another package for multi-state modeling that is designed for estimation of non-parametric and semi-parametric (i.e., Cox proportional hazards) models \citep{de2011mstate}. It can predict state probabilities and simulate multi-state trajectories (to facilitate prediction for clock-reset models) from cumulative hazard functions evaluated at specific times. \citet{williams2017cost} has adapted \pkg{mstate} for cost-effectiveness modeling applications. There are however, some disadvantages of using \pkg{mstate} for health economic modeling since it was not designed for it. \code{mssample()}---the core function for simulating multi-state models---does not natively support heterogeneous target populations, PSA, simultaneous simulation of multiple treatment strategies, or computation of QALYs and costs from the simulated output. It is also less computationally efficient than \pkg{hesim} for two main reasons. First, simulation of general cumulative hazards functions is considerably slower than simulation of parametric distributions when efficient random number generators are available. Second, it is written completely in \proglang{R} and incorporating PSA, multiple treatment strategies, or heterogeneous patients results in code that is not vectorized. To compare computational performance, we ran (see ``Benchmarks'' on the package website) 1,000 PSA iterations where 1 treatment strategy and 1,000 homogeneous patients were simulated through a 6-state Weibull model during each iteration. At the time of writing, the model completed in 2.8 seconds in \pkg{hesim} and 3.1 hours in \pkg{mstate}, showing that computational considerations are not just an academic concern.

\pkg{heemod} is the only other general purpose \proglang{R} package for cost-effectiveness modeling. As noted in the introduction, there are some important differences between \pkg{hesim} and \pkg{heemod} related to the supported model types and integration of statistical and economic modeling. In addition, as with \pkg{mstate}, simulations are not vectorized which could result in significantly reduced performance. To test this, we simulated a time-inhomogeneous total hip replacement example from \citet{briggs2006decision}. A cohort model in \pkg{heemod} completed in approximately 1.9 minutes while the same cohort model took approximately 1 second to run in \pkg{hesim} and an equivalent individual-level simulation in \pkg{hesim} ran in 7.4 seconds.

There are of course features in other \proglang{R} packages that are not in the current version of \pkg{hesim}. For example, \pkg{heemod} has an explicit function, \code{calibrate_model()}, for calibrating model parameters that are difficult to measure. Furthermore, tunnel states can be automatically created (rather than needing to be created manually) with \pkg{heemod} using the \code{state_time} variable so that costs or transition probabilities can very as a function of time in a health state. \pkg{hesim} is also limited in its support for value of information analyses. For instance, neither the expected value of partial perfect information (EVPPI) or the expected value of sample information (EVSI) can be computed with \pkg{hesim}, whereas the EVPPI can be computed with \pkg{BCEA}, \pkg{dampack}, and \pkg{SAVI} and EVSI can be computed with \pkg{EVSI} and \pkg{dampack}. In this respect, these packages are complementary and simulation output from \pkg{hesim} could be used as inputs to their more advanced value of information functions.

A number of extensions of the current version of \pkg{hesim} would be helpful. First, we plan to extend the individual-level simulation so that users can update input data during the simulation as a function of time elapsed and/or the health states transitioned to. This would allow potentially time-varying variables such as age to update each time a patient transitions to a new state. Second, we would like to add more examples to ensure that \pkg{hesim} is fit for purpose for a wide range of applications in health economics. Third, more complex models of utility \citep{kharroubi2007modelling, hernandez2013relationship} and costs \citep{nixon2005methods} could help better capture heterogeneity in preferences and spending. Fourth, it would be useful to add functionality that automatically creates tunnel states for users in cDTSTMs. Fifth, equity considerations could be incorporated by adding support for distributional CEA \citep{asaria2015distributional, asaria2016distributional, cookson2017using}. Sixth, functions could be added to convert \pkg{hesim} simulation output into formats required by other packages that can analyze the outputs of health economics models, such as those with more advanced value of information analysis capabilities. Finally, a larger aspiration is to provide support for parameterization via NMA, particularly for estimation of relative treatment effects in survival (e.g., for PSMs) and multi-state models. However, since existing \proglang{R} packages are lacking in this area, this could involve creation of a separate NMA package and a new dependency for \pkg{hesim}.
